Halving the cost of quantum addition

### 2017-09-12 paper.tex
Note that adder block is a variant of Cuccaro's, and improve the optimization condition figure
Applying more Dave advice - Noted the cost of initializing a phase gradient register (and generally rewrote that paragraph) - Standardized on 'Toffoli pair' instead of 'matched Toffoli' - Polished the conclusion

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &c_k &&\ctrl{2}&\qw     &\ctrl{3}&\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\ctrl{3}&\qw     &\ctrl{1}&\qw     &\qw &c_k     &&&&& \\
      &i_k &&\targ   &\ctrl{1}&\qw     &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw     &\ctrl{1}&\targ   &\ctrl{1}&\qw &i_k     &&&&& \\
      &t_k &&\targ   &\ctrl{1}&\qw     &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw     &\ctrl{1}&\qw     &\targ   &\qw &&&(t+i)_k &&& \\
      &    &&        &        &\targ   &\qw &    &c_{k+1}&    &    &    &\ldots &    &    &    &c_{k+1}&    &    &\targ   &\qw     &        &        &    &&&        &&& \\
    }
  }
  \caption{
	Adder circuit building-block with a T-count of 4.
  }
  \label{fig:full-adder-block}
\end{figure}

@@

For example, our improved adder allows the operation $R_Z(\theta)$ to be applied to $n$ qubits with a total T-cost $4n + O(\text{poly}(\lg \frac{1}{\epsilon}) \lg n)$ as follows.
First, iteratively apply our temporary adder to combine the target qubits into a single binary register storing their Hamming weight.
This takes $4n + O(\lg n)$ T gates to do.
Then, for each position $k$ in the Hamming weight register, synthesize and apply the operation $R_Z(\theta \cdot 2^k)$ to the register qubit at that position.
This uses a number of T gates proportional to some polynomial of the desired bit precision, times the number of bits in the output register.
The output register is exponentially smaller than $n$, so this cost is negligible for large $n$.
Finally, uncompute the Hamming weight register (using no T gates).
This applies the $n$ desired $R_Z(\theta)$ operations.

Another rotation-based operation that can be implemented via an adder is the $n$-bit phase gradient operation $\text{Grad}_n = \sum_{k=0}^{2^n-1} e^{2 i \pi k / 2^n} |k\rangle \langle k|$.
Normally this operation would be implemented by separately applying the operation $R_z(\pi 2^{-p})$ to each qubit of the target register, where $p$ is the qubit's index in the register and the number of T gates needed for each rotation depends on the maximum per-gate error $\epsilon$.
However, assuming a ``phase gradient register" prepared in the state $2^{-b/2} \sum_{k=0}^{2^b-1} e^{-2 i \pi k / 2^b} |k\rangle$ is available, the phase gradient operation can be performed via addition.
Add the target register into the phase gradient register, and phase kickback will apply the $\text{Grad}_n$ operation to the target.
Ignoring the one-time cost of initializing the reusable phase gradient register, this uses $4n + O(1)$ T gates.

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &&&\ustick{n}&\diagup\qw &\multigate{3}{U}&\qw     &\qw &&               &&&&\ustick{n}&\diagup\qw &\multigate{3}{U}&\qw     &\qw &&            &&\ustick{n}&\diagup\qw &\qw      &\multigate{3}{U}&\qw      &\qw     &\qw &\\
      &&&          &\ctrl{1}   &\ghost{U}       &\ctrl{1}&\qw &&               &&&&          &\ctrl{1}   &\ghost{U}       &\ctrl{1}&\qw &&            &&          &\ctrl{1}   &\qw      &\ghost{U}       &\qw      &\ctrl{1}&\qw &\\
      &&&          &\ctrl{1}   &\ghost{U}       &\ctrl{1}&\qw &&\stackrel{?}{=}&&&&          &\ctrl{2}   &\ghost{U}       &\ctrl{2}&\qw &&\Rightarrow &&          &\ctrl{2}   &\qw      &\ghost{U}       &\qw      &\ctrl{2}&\qw &\\
      &&&          &\targ      &\ghost{U}       &\targ   &\qw &&               &&&&          &\targ      &\ghost{U}       &\targ   &\qw &&            &&          &\qw        &\targ    &\ghost{U}       &\targ    &\qw     &\qw &\\
      &&&\qO       &\qw        &\qw             &\qw     &\qw &&               &&&&\qO       &\targ      &\qw             &\targ   &\qw &&            &&          &           &\ctrl{-1}&\qw             &\ctrl{-1}&\qw     &    &\\
    }
  }
  \caption{
	A sufficient condition for replacing a pair of Toffoli gates with a temporary logical-AND, saving 4 T gates.\\
	1) The later Toffoli gate must be uncomputing the earlier Toffoli gate.\\
	2) Intermediate operations must not be sensitive to the presence of the entangled ancilla.
  }
  \label{fig:pair-definition}
\end{figure}

For example, a simple way to construct Grover oracles is to translate a classical predicate into a reversible circuit made out of Toffoli gates.
This classical reversible circuit is then used to compute an output bit, so that a Z gate can be applied to the output bit before the circuit is run in reverse to uncompute the output.
Every Toffoli gate generated by this construction is paired (each Toffoli in the computation will match a Toffol in the uncomputation).
Using temporary logical-ANDs halves the T-count of this approach to constructing Grover oracles.

Matched Toffolis also appear in quantum circuits rooted in physics instead of abstract mathematics.
For example, we used temporary logical-ANDs to cut the projected T-count of a step in a simulation algorithm under development by nearly 30\% \citep{RyanEmails2017}.

@@

For over a decade, the T-count of addition has been $8n + O(1)$ \citep{Amy2013, Barenco1995, Cuccaro2004}.
In this paper we showed how to halve the leading factor of this cost.
Furthermore, we did so via a construction that applies to many other circuits: the temporary logical-AND gate.

Because the temporary logical-AND is so widely applicable, the T-count of many circuits is lower than previously thought.
We covered a few examples in this paper but, for brevity, we leave further low hanging fruit as an exercise for the reader.

It is interesting to consider if addition or temporary logical-AND gates can be done with still fewer T gates.
\autoref{fig:lower-bound-logical-AND} presents a simple proof that the T-count of the temporary logical-AND is at least 2.
Regardless of whether the true lower bound is 4, we expect that there are special cases with lower average T-counts.
For example, if a task involves repeating an action several times (e.g. counting by repeatedly conditionally incrementing a register), T gates from Toffolis in uncomputation work at the end of one repetition may cancel against T gates from Toffolis in computation work at the start of the next repetition.

@@

Regardless of whether there will be further T-count improvements for operations as basic as matched Toffoli gates, it is clear that many other T-count optimizations are waiting to be found.
Not just in the construction of basic low-level operations, but in medium-level constructions, high-level constructions, lower-than-circuit level constructions, and generally across the whole technology stack that will be needed to perform error corrected quantum computation with the surface code for the first time.

>>>>

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &c_k &&\ctrl{2}&\qw     &\ctrl{3}&\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\ctrl{3}&\qw     &\ctrl{1}&\qw     &\qw &c_k     &&&&& \\
      &i_k &&\targ   &\ctrl{1}&\qw     &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw     &\ctrl{1}&\targ   &\ctrl{1}&\qw &i_k     &&&&& \\
      &t_k &&\targ   &\ctrl{1}&\qw     &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw &\qw    &\qw &\qw &\qw     &\ctrl{1}&\qw     &\targ   &\qw &&&(t+i)_k &&& \\
      &    &&        &        &\targ   &\qw &    &c_{k+1}&    &    &    &\ldots &    &    &    &c_{k+1}&    &    &\targ   &\qw     &        &        &    &&&        &&& \\
    }
  }
  \caption{
	Adder circuit building-block with a T-count of 4.
	A variant of the Cuccaro adder building-block \citep{Cuccaro2004}.
  }
  \label{fig:full-adder-block}
\end{figure}

@@

For example, our improved adder allows the operation $R_Z(\theta)$ to be applied to $n$ qubits with a T-cost of $4n + O(\text{poly}(\lg \frac{1}{\epsilon}) \lg n)$ as follows.
First, iteratively apply our temporary adder to combine the target qubits into a single binary register storing their Hamming weight.
This takes $4n + O(\lg n)$ T gates to do.
Then, for each position $k$ in the Hamming weight register, synthesize and apply the operation $R_Z(\theta \cdot 2^k)$ to the register qubit at that position.
This uses a number of T gates proportional to some polynomial of the desired bit precision, times the number of bits in the output register.
The output register is exponentially smaller than $n$, so this cost is negligible for large $n$.
Finally, uncompute the Hamming weight register (using no T gates).
This applies the $n$ desired $R_Z(\theta)$ operations.

Another quantum operation that can be implemented via an adder is the $n$-qubit phase gradient operation $\text{Grad}_n = \sum_{k=0}^{2^n-1} e^{2 i \pi k / 2^n} |k\rangle \langle k|$.
Normally this operation would be implemented by separately applying the operation $R_z(\pi 2^{-p})$ to each qubit of the target register, where $p$ is the qubit's index in the register and the number of T gates needed for each rotation depends on the maximum per-gate error $\epsilon$.
However, assuming a ``phase gradient register" prepared in the state $2^{-b/2} \sum_{k=0}^{2^b-1} e^{-2 i \pi k / 2^b} |k\rangle$ is available, the phase gradient operation can be performed via addition.
Add the target register into the phase gradient register, and phase kickback will apply the $\text{Grad}_n$ operation to the target.
This construction performs phase gradients using $4n + O(1)$ T gates, which is apparently independent of $\epsilon$.
However, there are in fact three ways in which the construction's cost does depend on $\epsilon$.
First, $\epsilon$ still bounds the minimum quality of the $|A\rangle$ states powering the T gates.
Second, large phase gradients can be truncated down to a size $n_{\text{max}}$ asymptotically equal to $\lg \lg \epsilon$.
Third, initializing the reusable phase gradient register has a one-time cost of $O(n_{\text{max}} \cdot \text{poly}(\lg \frac{1}{\epsilon}))$ T gates.

@@

\begin{figure}
  \resizebox{\linewidth}{!}{
    \Qcircuit @R=0.7em @C=0.7em {
      &\ustick{n}&\diagup\qw &\multigate{3}{U}&\qw     &\qw &&               &&&&\ustick{n}&\diagup\qw &\multigate{3}{U}&\qw     &\qw &          &&            &&\ustick{n}&\diagup\qw &\qw      &\multigate{3}{U}&\qw      &\qw     &\qw &\\
      &          &\ctrl{1}   &\ghost{U}       &\ctrl{1}&\qw &&               &&&&          &\ctrl{1}   &\ghost{U}       &\ctrl{1}&\qw &          &&            &&          &\ctrl{1}   &\qw      &\ghost{U}       &\qw      &\ctrl{1}&\qw &\\
      &          &\ctrl{1}   &\ghost{U}       &\ctrl{1}&\qw &&\stackrel{?}{=}&&&&          &\ctrl{2}   &\ghost{U}       &\ctrl{2}&\qw &          &&\Rightarrow &&          &\ctrl{2}   &\qw      &\ghost{U}       &\qw      &\ctrl{2}&\qw &\\
      &          &\targ      &\ghost{U}       &\targ   &\qw &&               &&&&          &\targ      &\ghost{U}       &\targ   &\qw &          &&            &&          &\qw        &\targ    &\ghost{U}       &\targ    &\qw     &\qw &\\
      &          &           &                &        &    &&               &&&&\qO       &\targ      &\qw             &\targ   &\qw &|0\rangle &&            &&          &           &\ctrl{-1}&\qw             &\ctrl{-1}&\qw     &    &\\
    }
  }
  \caption{
	A sufficient condition for replacing a pair of Toffoli gates with a temporary logical-AND, saving 4 T gates.\\
	1) The later Toffoli gate must be uncomputing the earlier Toffoli gate.\\
	2) Intermediate operations must not be sensitive to the presence of the entangled ancilla.
  }
  \label{fig:pair-definition}
\end{figure}

For example, a simple way to construct Grover oracles is to translate a classical predicate into a reversible circuit made out of Toffoli gates.
This classical reversible circuit is then used to compute an output bit, so that a Z gate can be applied to the output bit before the circuit is run in reverse to uncompute the output.
Every Toffoli gate generated by this construction is part of a compute/uncompute pair (each Toffoli in the computation will match a Toffoli in the uncomputation).
Using temporary logical-ANDs halves the T-count of this approach to constructing Grover oracles.

Toffolis also appear in compute/uncompute pairs in quantum circuits rooted in physics instead of abstract mathematics.
For example, we used temporary logical-ANDs to cut the projected T-count of a step in a simulation algorithm under development by nearly 30\% \citep{RyanEmails2017}.

@@

For over a decade, the T-count of addition has been $8n + O(1)$ \citep{Amy2013, Barenco1995, Cuccaro2004}.
In this paper we showed how to halve the leading factor of this cost by replacing Toffolis in compute/uncompute pairs with temporary logical-ANDs.
The temporary logical-AND is a basic circuit construction widget that can be used in many circuits, and a good example of measurement breaking the symmetry between computation and uncomputation (allowing one to be more efficient than the other).
We covered a few examples of optimizing T-counts with our constructions but, for brevity, we leave further low hanging fruit as an exercise for the reader.

It is interesting to consider if addition or temporary logical-ANDs can be done with still fewer T gates.
\autoref{fig:lower-bound-logical-AND} presents a simple proof that the T-count of the temporary logical-AND is at least 2.
Although we suspect the true lower bound is 4, we also expect that there are special cases with lower T-counts.
For example, if a task involves repeating an action several times (e.g. counting by repeatedly conditionally incrementing a register), T gates from Toffolis in uncomputation work at the end of one repetition may cancel against T gates from Toffolis in computation work at the start of the next repetition.

@@

Regardless of whether there will be further T-count improvements for operations as basic as Toffoli gates appearing in pairs, it is clear that many other T-count optimizations are waiting to be found.
Not just in the construction of basic low-level operations, but in medium-level constructions, high-level constructions, lower-than-circuit level constructions, and generally across the whole technology stack that will be needed to perform error corrected quantum computation with the surface code for the first time.
